Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
4.3

\contractr is an R package that implements our type system for the R package
ecosystem. For functions with type annotations, it adds contracts for
type-checking arguments and results on function calls. On contract failure,
it reports a type mismatch warning with the package name, function name,
argument name, argument position, expected type signature, actual type
signature of the value observed, and a stack trace for the called
function. \contractr works by modifying function definition to insert a call
to the type-checking function. When the modified function is called, the
type-checking function is invoked on the modified function's arguments and
return value.

\contractr is implemented in 434 R LOC and 2358 C++ LOC. The core
implementation is in C++ to keep the runtime overhead low. It been designed and
tested with GNU R-3.5.0 but it also works with recent versions of R. It has been
hardened with a battery of 400 test cases. We have used it extensively during
the course of this work; firstly, for sanity checking of XXX type signatures
generated for the XXX packages during the development phase, and secondly, for
assessing the quality of these type signatures on XXX packages during the
evaluation phase. We discuss this further in Section~\ref{sec:evaluation}.

%
%
\subsubsection{Type Declarations}
Type declarations can be made available to \contractr in three ways: as
part of its internal database of types (which already has type signatures for
our corpus of 500 packages), as part of a designated \emph{TYPEDECLARATION} file
supplied by the package author and installed along with the package, or through
a user-level API function insert\_contract that lets the user insert contract to
a custom function with the type declaration provided as a string argument.

%
%
\subsubsection{Challenges}
Retrofitting a robust and efficient type-system as an external package in R
has been a significant undertaking. We had to contend with the complex
design choices of R to make it work without surprises for the end-users. We
describe two key challenges that we faced in the development of \contractr.

@@

\subsubsection{Usability}
With \contractr, we aim to provide a hassle-free path to the R developers
for adding types to untyped package code. Hence, we have paid a lot of
attention towards usability in its design. We discuss three such usability
features.

\contractr enables automatic type-checking on importing without any user
intervention. Just running \code{library(contractr)} R code is enough to
insert contracts, both, in packages pre-loaded in the user's workspace, and,
in packages that will be loaded eventually. To achieve this automation,
\contractr relies heavily on R's reflective and dynamic capabilities. On
loading, \contractr scans all the pre-loaded packages in the user's
workspace and inserts contracts in functions for which type signatures are
available. For all other packages installed on user's machine but not yet
loaded in the workspace, \contractr sets up package load hooks which are
executed by R when those packages are loaded. The hook registered by
\contractr insert contracts to those package's functions. Thus, developers
and users do not have to call the \contractr API to enable type-checking for
packages at any point of their programming workflow. Furthermore, \contractr
automatically removes contracts from all the package functions and restores
them to their original state when it is unloaded, an operation that is
rarely performed, but useful in interactive settings.

\contractr enables package authors to supply type declarations for their
package without requiring package code modifications. Type declarations can
be written alongside a \code{@type} section inside \roxygen function comment
blocks. \roxygen is an R package that enables authors to add documentation
to R functions in the form of organized plaintext sections which are
automatically exported to R's latex style custom documentation format.
\roxygen is a widely adopted package, used by over XXX R packages in
CRAN. It provides an extension API for adding custom documentation tag
sections in comment blocks. The sections are processed by \roxygen before
package build step and a registered method for each custom tag section is
invoked with the section data and documented object. \contractr uses this
mechanism to register a hook for a custom \code{@type} tag. This hook parses
the type declarations from the \code{@type} sections, extracts the function
name from the documented function definition, and stores them in a
TYPEDECLARATION file inside the package folder. This files is copied
verbatim on package installation and picked up \contractr when the package
is loaded. Here is an example of how a roxygen code block looks like with
our type declaration:

This feature enables a function and its type signature to coexist next to
each other, where they are more likely to remain synchronized. Furthermore,
the \contractr hook can also add the function's type declaration to its
documentation, which seamlessly integrate our type system with the existing
R tooling.

\contractr provides a very expressive API to the users, covering a variety
of uses cases. During interactive development, developers can explicitly
insert contracts by supplying the type declaration as a string argument to
the \code{insert\_contract} API function. Conversely, contracts can be
removed by calling the \code{remove\_contract} API on functions. Contracts
can be selectively enabled or disabled for code blocks by wrapping them in
calls to \code{ignore\_contracts} and \code{capture\_contracts}
functions. This enables selective type-checking of code section during the
development phase.  Furthermore, these functions return a data frame that
contains all the information about failed and successful contract assertions
in the wrapped code blocks. This is useful for post-hoc
investigation. Multiple type-checking failures can introduce a lot of noise
in the program output. To alleviate this problem, \code{contractr.severity}
option can be set to \code{'silence'}.  This suppresses failure messages but
\contractr still performs type-checking whose results can be explicitly
obtained as a data frame from the \code{get\_contracts} API function. R
suppresses printing of warnings on the terminal when a function issues too
many warnings. This can hide type-checking failure messages. Setting
\code{contractr.severity} flag to \code{'error'} turns the type-checking
failure warnings to errors, which halts the program at the point of contract
failure.

While \contractr's primary logic has been implemented in C++, we have not
observed a single segfault during our use of the package, either during the
development and sanity-checking phase, or during the evaluation phase.
We conclude the discussion of \contractr by drawing reader's attention to the
engineering effort it takes to package experimental ideas in the form of stable
and easy-to-use tools. R has an eclectic mix of features and design oddities
that make this even harder.

>>>>

One way to validate the inferred type annotations is to check that different
programs using one of the functions for which a type signature has been
inferred respect the signature.  \contractr is an R package that allows to
decorate function with assertions. We use it to insert type checks of the
arguments and return values. As usual with R, this is not straightforward.

While \contractr's primary logic has been implemented in C++ to reduce its
runtime overhead, we have not observed a single segfault during our use of
the package.  It has been tested with GNU R-3.5.0 and hardened with a
battery of 400 test cases. It works by modifying function definition to
insert a call to a type-checking function.

In terms of useability, our checker is enabled automatically whenever a
package is loaded in R. So, a simple invocation of \code{library(contractr)}
causes contracts to be injected in all packages.  On loading, \contractr
scans all packages in the user's workspace and inserts contracts in
functions for which type signatures are available. Then \contractr sets up
package load hooks which are executed when new packages are loaded.
Furthermore, \contractr automatically removes contracts from all functions
and restores them to their original state when it is unloaded.

The type signatures can be provided externally, in a file, thus avoiding the
need to change the source code of checked packages.  Type declarations can
also be written alongside a \code{@type} section inside \roxygen function
comment blocks. \roxygen is an R package that enables authors to add
documentation. The documentation sections are processed by \roxygen before
package build and a registered method for each custom tag section is
invoked with the section data and documented object. \contractr uses this
mechanism to register a hook for a custom \code{@type} tag.

The \contractr API allows users to explicitly insert contracts during
interactive development by supplying the type signature as a string.
Contracts can be removed by calling \code{remove\_contract}.  Contracts can
be selectively enabled or disabled for code blocks by wrapping them in calls
to \code{ignore\_contracts} and \code{capture\_contracts}.  These functions
return a data frame that contains all the information about failed and
successful contract assertions in the wrapped code blocks.  Multiple
failures can introduce noise; this is avoided by setting
\code{contractr.severity} to \code{'silence'}.  \contractr still performs
type-checking so that results can be obtained by calling
\code{get\_contracts}. Setting \code{contractr.severity} flag to
\code{'error'} turns the type-checking warnings to errors and
halt the program.

@@
